---
title: "Introduction to R pipes and dplyr"
author: "Paul M. Magwene"
output: beamer_presentation
fontsize: "9pt"    
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      warning=FALSE, 
                      message=FALSE, 
                      comment = "", 
                      collapse = TRUE)
options(width=70)
```

```{r, echo = FALSE}
library(tidyverse)
#library(palmerpenguins)
```


## The magrittr pipe operator

An R package called `magrittr` popularized a new operator called a pipe (Note: pipes are an older computing concept dating back to the origins of the Unix operating)

  - The pipe operator is `%>%`

Using pipes:

  - `x %>% f()` is equivalent to `f(x)`
  - `x %>% f(y)` is equivalent to `f(x,y)`.

Examples:

  - Single argument function:
    ```{r, echo=TRUE}
pi %>% cos()
```
  
  - For single argument functions, after the pipe you can drop the parentheses:
    ```{r, echo=TRUE}
pi %>% cos  # same as above
```

  - Multi-argument functions
    ```{r, echo=TRUE}
100 %>% log(base=10) # 100 is treated as the first argument
```


## Building pipelines with the pipe operator

The pipe operator allows us to build analysis "pipelines".

\bigskip

A pipeline series of function calls that filter and/or transform our data

\medskip

```{r, echo=TRUE}
letters %>%         # start with letters vector
  str_to_upper %>%  # convert to upper case
  tail(10) %>%      # get last 10 elements
  str_flatten("-")  # join into single string, separated by '-'
```

\bigskip

The pipe operator helps to make our intent clearer, as compared to nested function calls:

\medskip

```{r, echo=TRUE}
str_flatten(tail(str_to_upper(letters), 10), "-")
```

## The "built-in" pipe operator (R 4.1+)

The magrittr pipe operator became popular enough that the R developers added a native pipe operator to the R language starting in release 4.1.

- The native pipe operator is `|>`

\medskip

The native pipe operator works similarly to the magrittr pipe:

  - `x |> f()` is equivalent to `f(x)`
  - `x |> f(y)` is equivalent to `f(x,y)`.

\medskip

`magrittr` pipe allows you to drop parentheses for single argument functions, but this is not allowed with the native pipe:

\medskip

```{r, eval = FALSE}
pi %>% cos   # equivalent to pi %>% cos()
pi |> cos  # this is an error with the native pipe
pi |> cos()  # must add parentheses with native pipe
```  
  
## Placeholder, magrittr pipes

`magrittr` pipe allows you to use a "placeholder" symbol (`.`) for cases where the object your piping is not the first argument in the function being called:

```{r}
# the period (.) is the placeholder
c("A", "B", "C") %>% 
  data.frame(x = c(1,2,3), y = .)
```

\medskip

`magrittr` placeholder can appear multiple times:

```{r}
c("A", "B", "C") %>% 
  tibble(x = c(1,2,3), y = ., z = str_to_lower(.))
```

## Placeholder, native pipes

A placeholder symbol (`_`) for use with the native pipe was introduced in R 4.2, but it has several limitations:

- Can only appear once
- Can only be used with named arguments

\medskip

This works:

```{r}
c("A", "B", "C") |> 
  tibble(x = c(1,2,3), y = _)
```

\medskip

But this doesn't:

```{r, eval=FALSE}
c("A", "B", "C") |> 
  tibble(x = c(1,2,3), y = _, z = str_to_lower(_))
```



## What is `dplyr`?

`dplyr` is a package that provides a "grammar for data manipulation"

\bigskip

Core "verbs" in the `dplyr` package:

Column manipulation:

  * `select()`
  * `mutate()`  
  
Row manipulation:

  * `filter()`
  * `arrange()`
  
Collapsing and Grouping:

  * `summarize()`
  * `group_by()`
  
\bigskip

All these functions return new data frames instead of modifying existing data frames



## `select()` subsets columns

```{r, echo = TRUE}
# select two columns
select(iris, Sepal.Length, Petal.Length) |> head(3)
```

\medskip

```{r, echo = TRUE}
# select everything BUT the species column
select(iris, -Species) |> head(3)
```

## `select()` has some helper functions for powerful filtering

See https://tidyselect.r-lib.org/reference/language.html

\bigskip

* `var1:var10`: variables lying between var1 on the left and var10 on the right. Also works with numeric indices.
* `everything()`: all variables.
* `all_of(vars)/any_of(vars)`: matches names stored in the character vector vars. 
* `last_col()`: furthest column on the right.
* `where(predicate)`: all variables where predicate function returns TRUE. e.g `where(is.numeric)`
* `starts_with()`: Starts with an exact prefix.
* `ends_with()`: Ends with an exact suffix.
* `contains()`: Contains a literal string.
* `matches()`: Matches a regular expression.
* `num_range()`: Matches a numerical range like x01, x02, x03.


## `select()` helper examples

```{r, size = "small"}
select(iris, starts_with("Petal")) |> head(3)
```

\medskip

```{r, size = "small"}
select(iris, ends_with("Length")) |> head(3)
```

\medskip
```{r, size = "small"}
select(iris, where(is.numeric)) |> head(3)
```


## `mutate()` adds or transforms columns

\footnotesize

```{r, echo=TRUE}
mutate(iris, Species = str_to_upper(Species)) |> head(3)
```

\medskip

```{r}
# mutate can refer to multiple existing columns
iris |>
  select(starts_with("Sepal")) |>
  mutate(Sepal_Sum = Sepal.Width + Sepal.Length) |>
  head(3)
```

\medskip

```{r}
# can refer to columns created "on the fly"
iris |>
  select(starts_with("Sepal")) |>
  mutate(Sepal_Sum = Sepal.Width + Sepal.Length,
         Sepal_Width_Ratio = Sepal.Width / Sepal_Sum) |>
  head(3)
```

\normalsize


## `filter()` selects rows that match criteria

```{r, echo = TRUE}
# get only the I. setosa specimens
filter(iris, Species == "setosa") |> head(3)
```

\medskip

```{r, echo=TRUE}
# filter on multiple criteria, treated as "AND"
filter(iris, Species == "setosa", Sepal.Length < 5) |> head(3)
```

## `arrange()` sorts rows according to values of one or more columns

\small

```{r, size="small"}
# sort by Sepal.Length
arrange(iris, Sepal.Length) |> head(3)
```

\medskip

```{r, size="small"}
# sort on multiple columns: by Sepal.Length then by Petal.Length
arrange(iris, Sepal.Length, Petal.Length) |> head(3)
```

\medskip

```{r, size = "small"}
# sort in descending order using helper desc()
arrange(iris, desc(Sepal.Length)) |> head(3)
```

\normalsize


## `summarize()` transforms and collapses

`summarize()` applies functions to one or more variables (columns) in the data frame, reducing a vector of values to a single value and returning the results in a data frame

\bigskip

```{r, echo=TRUE, collapse_vspace=TRUE}
summarize(iris, 
          avg.Sepal.Length = mean(Sepal.Length), 
          avg.Petal.Length = mean(Petal.Length))

```

## `group_by()` is used for conditioning (faceting) and transforming

```{r, echo=TRUE}
# apply grouping
grouped.df <- group_by(iris, Species)

# summarize grouped data frame
summarize(grouped.df,
          avg.Sepal.Length = mean(Sepal.Length),
          avg.Petal.Length = mean(Petal.Length))

```



## The `dplyr` functions are designed to work well with piping!


```{r}
iris |>
  filter(Species != "virginica") |>
  group_by(Species) |>
  summarize(avg.Sepal.Length = mean(Sepal.Length),
            avg.Sepal.Width = mean(Sepal.Width))   
```

## dplyr pipelines feed naturally into ggplot


But note that `ggplot` layers are "added" *not* piped:

\medskip

\footnotesize
```{r, collapse=FALSE, fig.align='center',out.width = "60%"}
iris |>
  filter(Species != "virginica") |>
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + 
    geom_point(size=3) + 
    labs(x = "Sepal Length", y = "Sepal Width",
         title = "Sepal Morphology for non-virginica Iris species") +
    theme_classic() +
    theme(legend.justification = c(1, 1), legend.position = c(1, 1))
```
\normalsize

## Other `dplyr` verbs of interest

Column manipulation:

* `rename()` - change column names
* `relocate()` - change column positions
* `pull()` - similar to `$`, returns a vector rather than a data frame

\medskip

Row manipulation:

* `slice()` - index rows by integer locations.
  - `slice` variants: `slice_head()`, `slice_tail()`, `slice_min()`, `slice_max()`, `slice_sample()`
* `distinct()` - collapse data frames into rows that are unique; usually used with subset of columns

Group manipulation

* `count()` - counts observations in a specified groups

