---
title: "Bio 723, Assignment 01"
author: "Your Name and NetID here!"
output:
  html_document:
    highlight: default
    theme: readable
    fig_caption: true
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, prompt = FALSE, eval = FALSE, 
                      warning = TRUE, comment=NA, cache = FALSE,
                      fig.width = 6, fig.height = 4)
```


# Instructions

Create a R Markdown document and answer the questions below using code blocks that generate the correct outputs. Feel free to also use explanatory text in your markdown document, however **each of your solutions should show how to generate the answer using R code**. 

Make sure your R Markdown is nicely formatted -- use headers, bullets, numbering, etc so that the structure of the document is clear.

Make sure to "knit" you document before submission to confirm that all code blocks and formatting is as you expect.

When completed, title your R Markdown file as follows:

- `netid-assignment_01-Spring2023.Rmd`

Submit both your Markdown file and the generated HTML document via the Assignments submission section on Sakai.



# 1. Numeric data

 a. What is the type of the numeric value `2`? [0.25 pt]

 a. What is the type of the numeric value `2L`? [0.25 pt]

 a. What is the value of $1/0$? What is its type? [0.25 pt]

 a. What is the value of $0/0$? What is its type? [0.25 pt]

 a. What is the output of `sqrt(-1)` ($\sqrt{-1}$)?  What is its type?  [0.25 pt]
 
 a. What is the value of `sqrt(-1 + 0i)`? What is its type?  Why does this differ from the previous result? [0.25 pt]

# 2. Character data 

a. Evaluate the following code. Does it give the output you expect? Why or why not? Can you identify a function that returns the number of characters in a string? [1 pt]
 
    ```{r}
    x <- "Hello, World!"
    length(x)
    ```

a. `paste` and `stringr::str_c` both join strings, but given the same inputs they produce different outputs.  Modify the call to `str_c` below to make the outputs identical. [1 pt]
  
    ```{r}
    x <- "hello"
    y <- "world"
    paste(x, y)
    str_c(x, y)
    ```

a. What does the function `stringr::str_to_lower` do?  Give a code block illustrating the use of this function. [1 pt]
  
a. What does the function `stringr::str_split` do? Give a code block illustrating the use of this function? [1 pt]
  

# 3. Vectors

a. `letters` is a pre-defined vector of characters giving the lowercase letters of the English alphabet.
 
    ```{r}
    letters
    ```
 
    Write a code block that combines indexing of `letters` with the `seq` function to retrieve every 3rd letter of the alphabet.  Your output should look like this [2 pts]:
 
      ```
      [1] "a" "d" "g" "j" "m" "p" "s" "v" "y"
      ```
    
a. This is a multi-part problem:
    - Create a vector, $x$, with all the even numbers between 2 and 100, inclusive. Do **NOT** generate this by hand! [1 pt]  
    - Show how to use vector indexing to extract all the numbers from $x$ that are evenly divisible by 3 (HINT: the modulo operator, `%%`, will be useful for this [1 pt]
    - Show how to replace all the elements of $x$ that are greater than 20 and less than 80 with the value -99. [1 pt]

a. The following two vectors give names of countries and gross domestic product (GDP) for each of those countries (expressed in USD):
    ```{r}
    countries <- c("Benin", "Nepal", "United Arab Emirates", 
      "United States", "Solomon Islands", "Ireland", "Qatar", 
      "Macau", "Singapore", "Norway", "Luxembourg", "Brunei", 
      "Switzerland", "Hong Kong", "Senegal", "Yemen")
    
    gdp <- c(2168, 2468, 72419, 57467, 2236, 68883, 
      127523, 104169, 87856, 59302, 105882, 77441, 
      62882, 58553, 2568, 2508)
    ```
   
    - Write a code block showing how you could find all the countries with a per capita GDP greater than $70,000. [1 pt]
  
    - Write a code block showing how to show sort the `gdp` vector in descending order (i.e. from largest to smallest) [1 pt]
  
    - Write a code block showing how to  sort the `countries` vector from with respect to GDP, with the countries with the largest GDPs listed first. [2 pts]    
  
   
 
# 4. Lists


Here's code that shows how we could use lists to make a simple phone book.

```{r}
john <- list(fname = "John", lname = "Smithee", age = 24,
             phone = "919-555-1234")
jane <- list(fname = "Jane", lname = "Doe", age = 39,
             phone = "919-555-4321")
jack <- list(fname = "Jack", lname = "Indabox", age = 13,
             phone = NA)
jill <- list(fname = "Jill", lname = "Bopeep", age = 18,
             phone = "919-555-1243")

phone.book <- list(john, jane, jack, jill)
```

 a. Write a code block illustrating how you could use list indexing to get back the first and last name of the first person in the phone book.  The returned object should be a list. [1 pt]

 a. Write a code block illustrating how you could use list indexing to look up the age of the fourth person in the phone book. The returned object should be a number. [1 pt]
 
 a. Write a code block illustrating how you could add a person to the `phone.book` list object. [1 pt]
 
 
# 5. Data frames

Here's an examplar data frame

```{r}
age <- c(30, 26, 21, 29, 25, 22, 28, 24, 23, 20)
sex <- rep(c("M","F"), 5)
wt.in.kg <- c(88, 76, 67, 66, 56, 74, 71, 60, 52, 72)

df <- data.frame(age = age, sex = sex, wt = wt.in.kg)
```

 a. Write a code block illustrating how to add a new column, "wt.in.lbs", giving the weight in US pounds (1 kg = 2.2 lbs) for every individual in `df` [1 pt]

 a. Write a code block showing how to use indexing to extract only the female individuals from `df` [1 pt]
 

 a. Write a code block showing how to create a new data frame that includes only the odd numbered rows (i.e. rows 1, 3, 5, ...) and just the age and sex columns from `df` [1 pt]